---
title: "Numbers used in text of the manuscript"
author: "Mathew Kiang"
date: "`r Sys.Date()`"
output: 
  html_document:
    code_folding: 'hide'
    toc: true
    toc_float: true
---

```{r options, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(error = FALSE)
```

## Set up

```{r settings, results='hide'}
library(tidyverse)
library(here)
library(knitr)
library(kableExtra)
source(here("codes", "utils.R"))
```

```{r data}
# Incidence of bereavement of at least on parent
# by cause and number of orphans
df.inc.parental.loss.by.cause <- readRDS(
        here(
                "outputs",
                "data_public",
                "df_incidence_parental_loss_by_cause.rds")
)

# Children experiencing bereavement of at least on parent
# by cause 
df.n.loss.by.cause <- readRDS(
        here(
                "outputs",
                "data_public",
                "df_n_loss_by_cause.rds")
)

# Life table
lt.US <- readRDS(
        here("outputs",
             "data_public",
             "lt_US.rds")
)

# Incidence of bereavement of at least on parent by cause and number of orphans
df.inc.parental.loss.by.cause.sex <- readRDS(
        here(
                "outputs",
                "data_public",
                "df_incidence_parental_loss_by_cause_sex.rds")
)

# Children experiencing bereavement of at least on parent by cause 
df.n.loss.by.cause.sex <- readRDS(
        here(
                "outputs",
                "data_public",
                "df_n_loss_by_cause_sex.rds")
)

```

## Demographics of drug and firearm deaths

### Number and percents of drug and firearm deaths by sex and cause of death
```{r}
lt.US |> 
    bind_rows(lt.US |> mutate(sex = "both")) |>
    mutate(drugs_and_firearms = n_drug + n_firearm) |> 
    filter(year %in% 1999:2020, race_eth == "total") |> 
    group_by(sex) |>
    summarize(n_drugs = sum(n_drug, na.rm = TRUE),
              n_firearm = sum(n_firearm, na.rm = TRUE),
              drugs_and_firearms = sum(drugs_and_firearms, na.rm = TRUE)) |> 
    mutate(percent_total = drugs_and_firearms / max(drugs_and_firearms) * 100) |> 
    mutate(across(n_drugs:drugs_and_firearms, ~ prettyNum(.x, big.mark = ",")),
           percent_total = sprintf("%0.1f", round(percent_total, 1)))
```

### Mean (SD) age by sex and type of death
```{r}
w_mean_age <- lt.US |>
    mutate(drugs_and_firearms = n_drug + n_firearm) |>
    filter(year %in% 1999:2020, race_eth == "total") |>
    summarize(weighted_mean = weighted.mean(age, drugs_and_firearms)) |> 
    pull(weighted_mean)
age_vector <- lt.US |>
    mutate(drugs_and_firearms = n_drug + n_firearm) |>
    filter(year %in% 1999:2020, race_eth == "total") |>
    pull(age)
weight_vector <- lt.US |>
    mutate(drugs_and_firearms = n_drug + n_firearm) |>
    filter(year %in% 1999:2020, race_eth == "total") |>
    pull(drugs_and_firearms)
w_sd_age <- sqrt(weighted.mean((age_vector - w_mean_age)^2, weight_vector))
print(sprintf("mean (SD) age of death of %0.1f (%0.1f) years", 
              round(w_mean_age, 1), 
              round(w_sd_age, 1)))
```

### Number and percent of drug and firearm deaths by race/ethnicity

For completeness, I include American Indian / Alaska Native and Asian / Pacific Islander but they were not analyzed in the manuscript (due to small sample size). 
```{r}
lt.US |>
    mutate(drugs_and_firearms = n_drug + n_firearm) |>
    filter(year %in% 1999:2020) |>
    group_by(race_eth) |>
    summarize(
        n_drugs = sum(n_drug, na.rm = TRUE),
        n_firearm = sum(n_firearm, na.rm = TRUE),
        drugs_and_firearms = sum(drugs_and_firearms, na.rm = TRUE)
    ) |>
    rename(race = race_eth) |> 
    mutate(percent_total = drugs_and_firearms / max(drugs_and_firearms) * 100) |>
    mutate(across(n_drugs:drugs_and_firearms, ~ prettyNum(.x, big.mark = ",")),
           percent_total = sprintf("%0.1f", round(percent_total, 1))) |> 
    categorize_race() |> 
    arrange(race_cat) |> 
    select(race_cat, race, n_drugs:percent_total)
```

## Number of youth experiencing parental deaths

### Number of parental deaths by race/ethnicity for each cause separately
```{r}
df.n.loss.by.cause |>
    categorize_race() |>
    ungroup() |>
    summarise(.by = c(cause, race_cat),
              across(median:upper95, ~ sum(.x))) |>
    arrange(race_cat, cause) |>
    transmute(race_cat,
              cause,
              estimate = sprintf(
                  "%0.1f (95%% CI: %0.1f to %0.1f)",
                  round(median  / 1000, 1),
                  round(lower95 / 1000, 1),
                  round(upper95 / 1000, 1)
              )) |>
    arrange(cause, race_cat) |> 
    kable(
        row.names = FALSE,
        col.names = c(
            "Race/Ethnicity",
            "Cause of death",
            "Estimate in thousands (95% CI)"
        ),
        format = "html",
        booktabs = TRUE
    ) |>
    kableExtra::kable_styling(bootstrap = c("striped", "scale_down"))
```

### Number of losses due to drugs and firearms combined by race/ethnicity
```{r}
df.n.loss.by.cause |>
    filter(cause != "other") |> 
    ungroup() |>
    summarise(.by = c(race),
              across(median:upper95, ~ sum(.x))) |>
    mutate(cause = "drugs and firearms") |>
    arrange(race, cause) |>
    categorize_race() |> 
    transmute(race_cat,
              cause,
              estimate = sprintf(
                  "%s (95%% CI: %s to %s)",
                  prettyNum(round(median  / 1000, 1), big.mark = ","),
                  prettyNum(round(lower95 / 1000, 1), big.mark = ","),
                  prettyNum(round(upper95 / 1000, 1), big.mark = ",")
              )) |>
    arrange(race_cat) |> 
    kable(
        row.names = FALSE,
        col.names = c(
            "Race/Ethnicity",
            "Cause of death",
            "Estimate in thousands (95% CI)"
        ),
        format = "html",
        booktabs = TRUE
    ) |>
    kableExtra::kable_styling(bootstrap = c("striped", "scale_down"))
```

## Per-thousand of parental deaths 

### Per-thousand of losses due to both drugs and firearms in 2020 by race/ethnicity

```{r}
df.inc.parental.loss.by.cause |>
    filter(cause != "other",
           year == 2020) |>
    ungroup() |>
    summarise(.by = c(race),
              across(median:upper95, ~ sum(.x) * 1000)) |>
    arrange(race) |>
    mutate(cause = "drugs and firearms") |>
    transmute(race,
              cause,
              estimate = sprintf(
                  "%0.2f per 1000 (95%% CI: %0.2f to %0.2f)",
                  round(median, 2),
                  round(lower95, 2),
                  round(upper95, 2)
              )) |>
    kable(
        row.names = FALSE,
        col.names = c("Race/Ethnicity",
                      "Cause of death",
                      "Estimate per 1000 (95% CI)"),
        format = "html",
        booktabs = TRUE
    ) |>
    kableExtra::kable_styling(bootstrap = c("striped", "scale_down"))
```

### Per-thousand of children experiencing parental death in 2020 by race/ethnicity and cause of death

```{r}
df.inc.parental.loss.by.cause |>
    categorize_race() |>
    ungroup() |> 
    filter(year == 2020) |>
    mutate(across(median:upper95, ~ .x * 1000)) |>
    transmute(
        race_cat,
        cause,
        year,
        estimate = sprintf(
            "%0.2f per 1000 (95%% CI: %0.2f to %0.2f)",
            round(median,  2),
            round(lower95, 2),
            round(upper95, 2)
        )
    ) |>
    arrange(desc(cause), race_cat) |>
    kable(
        row.names = FALSE,
        col.names = c(
            "Race/Ethnicity",
            "Cause of death",
            "Year",
            "Estimate per 1000 (95% CI)"
        ),
        format = "html",
        booktabs = TRUE
    ) |>
    kableExtra::kable_styling(bootstrap = c("striped", "scale_down"))
```
## Proportion of parentals deaths

### Proportion of all parental deaths due to drugs and firearms over whole time period
```{r}
prop_deaths <- df.n.loss.by.cause |>
    categorize_race() |>
    ungroup() |>
    summarise(.by = c(cause, race_cat),
              across(median:upper95, ~ sum(.x))) |>
    arrange(race_cat, cause) |> 
    filter(race_cat == "Total") |> 
    mutate(prop = median / sum(median)) |> 
    filter(cause != "other") |> 
    pull(prop) |> 
    sum()
print(sprintf("drugs and firearms account for %0.1f%% of all paternal deaths between 1999 and 2020", prop_deaths * 100))
```

### Proportion of all parental deaths due to drugs and firearms in 1999
```{r}
prop_deaths <- df.n.loss.by.cause |>
    filter(year == 1999) |> 
    categorize_race() |>
    ungroup() |>
    summarise(.by = c(cause, race_cat),
              across(median:upper95, ~ sum(.x))) |>
    arrange(race_cat, cause) |> 
    filter(race_cat == "Total") |> 
    mutate(prop = median / sum(median)) |> 
    filter(cause != "other") |> 
    pull(prop) |> 
    sum()
print(sprintf("drugs and firearms account for %0.1f%% of all paternal deaths in 1999", prop_deaths * 100))
```

### Proportion of all parental deaths due to drugs and firearms in 2020
```{r}
prop_deaths <- df.n.loss.by.cause |>
    filter(year == 2020) |> 
    categorize_race() |>
    ungroup() |>
    summarise(.by = c(cause, race_cat),
              across(median:upper95, ~ sum(.x))) |>
    arrange(race_cat, cause) |> 
    filter(race_cat == "Total") |> 
    mutate(prop = median / sum(median)) |> 
    filter(cause != "other") |> 
    pull(prop) |> 
    sum()
print(sprintf("drugs and firearms account for %0.1f%% of all paternal deaths in 2020", prop_deaths * 100))
```

## Relative changes 

### Relative increase in losses from 1999 to 2020 for total population by cause of death

```{r}
df.n.loss.by.cause |>
    filter(race == "total",
           year %in% c(1999, 2020)) |>
    ungroup() |>
    summarise(.by = c(race, cause, year),
              across(median:upper95, ~ sum(.x))) |>
    pivot_wider(names_from = year,
                values_from = median:upper95) |>
    mutate(
        m_perc_inc = ((median_2020 - median_1999) / median_1999) * 100,
        u95_perc_inc = ((lower95_2020 - lower95_1999) / lower95_1999) * 100,
        l95_perc_inc = ((upper95_2020 - upper95_1999) / upper95_1999) * 100
    ) |>
        transmute(race, 
                  cause,
                  estimate = sprintf("%0.1f%% (95%% CI: %0.1f to %0.1f)", 
                                     round(m_perc_inc, 1),
                                     round(l95_perc_inc, 1),
                                     round(u95_perc_inc, 1))) |> 
        kable(
            row.names = FALSE, 
            col.names = c("Race/Ethnicity",
                          "Cause of death",
                           "Estimated change in % (95% CI)"),
            format = "html",
            booktabs = TRUE
        ) |> 
    kableExtra::kable_styling(bootstrap = c("striped", "scale_down"))
```

## Parent-specific results

### Per-thousand of losses due to both drugs and firearms in 2020 by race/ethnicity and parent

```{r}
df.inc.parental.loss.by.cause.sex |>
    filter(cause != "other",
           year == 2020) |>
    ungroup() |>
    summarise(.by = c(race, sex),
              across(median:upper95, ~ sum(.x) * 1000)) |>
    arrange(sex, race) |>
    mutate(cause = "drugs and firearms") |>
    transmute(sex,
              race,
              cause,
              estimate = sprintf(
                  "%0.2f per thousand (95%% CI: %0.2f to %0.2f)",
                  round(median,  2),
                  round(lower95, 2),
                  round(upper95, 2)
              )) |>
    kable(
        row.names = FALSE,
        col.names = c(
            "Parent",
            "Race/Ethnicity",
            "Cause of death",
            "Estimate in per 1000 (95% CI)"
        ),
        format = "html",
        booktabs = TRUE
    ) |>
    kableExtra::kable_styling(bootstrap = c("striped", "scale_down"))
```

### Per-thousand of children experiencing parental death in 2020 by race/ethnicity and cause of death and parent

```{r}
df.inc.parental.loss.by.cause.sex |>
    categorize_race() |>
    ungroup() |> 
    filter(year == 2020) |>
    mutate(across(median:upper95, ~ .x * 1000)) |>
    transmute(
        sex, 
        race_cat,
        cause,
        year,
        estimate = sprintf(
            "%0.2f per 1000 (95%% CI: %0.2f to %0.2f)",
            round(median,  2),
            round(lower95, 2),
            round(upper95, 2)
        )
    ) |>
    arrange(sex, desc(cause), race_cat) |>
    kable(
        row.names = FALSE,
        col.names = c(
            "Parent", 
            "Race/Ethnicity",
            "Cause of death",
            "Year",
            "Estimate per 1000 (95% CI)"
        ),
        format = "html",
        booktabs = TRUE
    ) |>
    kableExtra::kable_styling(bootstrap = c("striped", "scale_down"))
```



## Other numbers (some not used in the text, just for comparison)

### Increase in losses due to firearms for black between 2014 and 2020

```{r}
df.inc.parental.loss.by.cause |> 
        filter(
                cause == "firearms",
                race == "black",
                year %in% c(2014,2020)
        ) |> 
        pivot_wider(
                names_from = year, 
                values_from = median:upper95
        ) |> 
        mutate(
                m_perc_inc = ((median_2020 - median_2014) / median_2014) * 100,
                u95_perc_inc = ((lower95_2020 - lower95_2014) / lower95_2014) * 100,
                l95_perc_inc = ((upper95_2020 - upper95_2014) / upper95_2014) * 100
        ) |> 
        dplyr::select(
                cause, race, contains("perc_inc")
        ) |>
    transmute(race,
              cause,
              estimate = sprintf(
                  "%0.2f (%0.2f to %0.2f)",
                  round(m_perc_inc, 2),
                  round(l95_perc_inc, 2),
                  round(u95_perc_inc, 2)
              )) |>
    kable(
        row.names = FALSE,
        col.names = c("Race/Ethnicity",
                      "Cause of death",
                      "Estimated change in % (95% CI)"),
        format = "html",
        booktabs = TRUE
    ) |>
    kableExtra::kable_styling(bootstrap = c("striped", "scale_down"))
```

### Number of parental deaths due to *all* causes combined by race/ethnicity
```{r}
df.n.loss.by.cause |>
    categorize_race() |>
    ungroup() |>
    summarise(.by = c(race_cat),
              across(median:upper95, ~ sum(.x))) |>
    arrange(race_cat) |>
    transmute(race_cat,
              estimate = sprintf(
                  "%0.1f (%0.1f to %0.1f)",
                  round(median  / 1000, 1),
                  round(lower95 / 1000, 1),
                  round(upper95 / 1000, 1)
              )) |>
    arrange(race_cat) |> 
    kable(
        row.names = FALSE,
        col.names = c(
            "Race/Ethnicity",
            "Estimate in thousands (95% CI)"
        ),
        format = "html",
        booktabs = TRUE
    ) |>
    kableExtra::kable_styling(bootstrap = c("striped", "scale_down"))
```
